Modelling M/EEG data with Bayesian generalised additive multilevel models

### 2025-02-13 brms_meeg.tex
pushing some preliminary simulation results

\title{Modelling M/EEG data with Bayesian multilevel generalised
additive models}

@@

\abstract{Time-resolved electrophysiological measurements such as those
offered by magneto- or electro-encephalography (M/EEG) provide a unique
window onto neural activity underlying cognitive process and how they
unfold over time. Typically, we are interested in testing whether such
measures differ across conditions and/or groups. The conventional
approach consists in conducting mass-univariate statistics through
followed by some form of multiplicity correction (e.g., FDR, FWER) or
cluster-based inference. However, these cluster-based methods have an
important downside: they shift the focus of inference from the timepoint
to the cluster level, thus preventing any conclusion to be made about
the onset and offset of effects (e.g., differences across conditions).
Here, we introduce a novel \emph{model-based approch} for analysing
one-dimensional M/EEG timeseries such as ERPs or decoding timecourses
and their differences across conditions or group. This approach relies
on Bayesian nonparametric multilevel modelling (multilevel generalised
additive models or Gaussian processes), which outputs the posterior
probabilility of the effect being different than 0 (or above chance) at
every timestep, while naturally taking into account the temporal
dependencies and between-subject variability present in such data.}

\keywords{EEG, MEG, generalised additive models, gaussian
processes, mixed-effects models, Bayesian statistics, brms}

@@

\hypertarget{toc}{}
\tableofcontents
\newpage
\section[Introduction]{Modelling M/EEG data with Bayesian multilevel
generalised additive models}

@@

\section{Introduction (in progress)}\label{introduction-in-progress}

Here are some useful references to be discussed
(\citeproc{ref-combrisson_exceeding_2015}{Combrisson \& Jerbi, 2015};
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-frossard2021}{Frossard \& Renaud, 2021},
\citeproc{ref-frossard2022}{2022}; \citeproc{ref-gramfort2013}{Gramfort,
2013}; \citeproc{ref-hayasaka_validating_2003}{Hayasaka, 2003};
\citeproc{ref-luck_how_2017}{Luck \& Gaspelin, 2017};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al., 2019};
\citeproc{ref-pernet2015}{C. R. Pernet et al., 2015};
\citeproc{ref-riutort-mayol_practical_2023}{Riutort-Mayol et al., 2023};
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025})\ldots{} See also
(\citeproc{ref-maris2011}{Maris, 2011})\ldots{} and
(\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}) (history of
cluster-based approaches and using a data split?)\ldots{} Cluster
failure (\citeproc{ref-eklund2016}{Eklund et al., 2016})\ldots{}

In the following, we consider two approaches to modelling the non-linear
timecourse of M/EEG ERPs or decoding performance: i) generalised
additive models (GAMs) with thin-plate smoothing splines
(\citeproc{ref-wood2003}{Wood, 2003}; \citeproc{ref-wood2004}{Wood,
2004}) and ii) Gaussian processes (GPs) with a smooth covariance kernel
(\citeproc{ref-rasmussen2005}{Rasmussen \& Williams, 2005}) and low-rank
approximation (\citeproc{ref-riutort-mayol_practical_2023}{Riutort-Mayol
et al., 2023}).

@@

From Dimigen \& Ehinger (\citeproc{ref-dimigen2021}{2021}): ``Recently,
spline regression has been applied to ERPs (Hendrix, Baayen, \& Bolger,
2017; Kryuchkova,Tucker,Wurm,\&Baayen,2012;Tremblay\& Baayen, 2010;
Tremblay \& Newman, 2015)''\ldots{}

Disentangling overlapping processes
(\citeproc{ref-skukies_brain_2024}{Skukies et al., 2024};
\citeproc{ref-skukies_modelling_2021}{Skukies \& Ehinger, 2021})\ldots{}
Using Bayes factors (\citeproc{ref-teichmann2022}{Teichmann,
2022})\ldots{} Weighting single trials (\citeproc{ref-pernet2022}{C.
Pernet, 2022})\ldots{}

See
\url{https://www.let.rug.nl/nerbonne/teach/rema-stats-meth-seminar/presentations/Wieling-2014-GAMs.pdf}
and \url{https://www.let.rug.nl/wieling/Statistics/GAM-EEG/lab/},
\url{https://www.let.rug.nl/wieling/Statistics/GAM-EEG/GAM-EEG.pdf}
(\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-meulman2015}{Meulman et al., 2015})\ldots{}

@@

@@

@@

@@

We then fitted a GAM and GP regression model using the \texttt{brms}
package (\citeproc{ref-brms2017}{Bürkner, 2017},
\citeproc{ref-brms2018}{2018}; \citeproc{ref-nalborczyk2019}{Nalborczyk
et al., 2019}). We used the default priors in \texttt{brms}, that is,
weakly informative priors. Four Markov Chain Monte-Carlo (MCMC) were ran
for each model to approximate the posterior distribution, including each
5000 iterations and a warmup of 2000 iterations. Posterior convergence
was assessed examining trace plots as well as the Gelman--Rubin statisti
\(\hat{R}\).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# averaging across participants}
\NormalTok{ppt\_df }\OtherTok{\textless{}{-}}\NormalTok{ raw\_df }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{group\_by}\NormalTok{(participant, condition, time) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{eeg =} \FunctionTok{mean}\NormalTok{(eeg) ) }\SpecialCharTok{\%\textgreater{}\%}
    \CommentTok{\# mutate(eeg = as.numeric(scale(x = eeg, scale = FALSE) ) ) \%\textgreater{}\%}
    \FunctionTok{ungroup}\NormalTok{()}

@@

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

Now we fit the GP regression model\ldots{}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# computing eeg\_diff and averaging across participants}
\CommentTok{\# group\_df \textless{}{-} raw\_df \%\textgreater{}\%}
\CommentTok{\#     group\_by(participant, time) \%\textgreater{}\%}
\CommentTok{\#     pivot\_wider(names\_from = condition, values\_from = eeg, values\_fn = mean) \%\textgreater{}\%}
\CommentTok{\#     mutate(eeg\_diff = cond2 {-} cond1) \%\textgreater{}\%}
\CommentTok{\#     summarise(eeg\_diff = mean(eeg\_diff) ) \%\textgreater{}\%}
\CommentTok{\#     ungroup()}

\CommentTok{\# see https://betanalpha.github.io/assets/case\_studies/gp\_part3/part3.html}
\CommentTok{\# and https://betanalpha.github.io/assets/case\_studies/gaussian\_processes.html}
\NormalTok{gp\_priors }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}
    \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"normal(0, 1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"Intercept"}\NormalTok{),}
    \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"normal(0, 1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"b"}\NormalTok{),}
    \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"exponential(1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"sigma"}\NormalTok{),}
    \FunctionTok{set\_prior}\NormalTok{(}\StringTok{"exponential(1)"}\NormalTok{, }\AttributeTok{class =} \StringTok{"sdgp"}\NormalTok{)}
\NormalTok{    )}

\NormalTok{gp\_model }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# k refers to the number of basis functions for}
    \CommentTok{\# computing Hilbert{-}space approximate GPs}
    \CommentTok{\# if k = NA (default), exact GPs are computed}
    \CommentTok{\# eeg \textasciitilde{} gp(time, by = condition),}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{gp}\NormalTok{(time, }\AttributeTok{by =}\NormalTok{ condition, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{cov =} \StringTok{"exp\_quad"}\NormalTok{),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \CommentTok{\# prior = gp\_priors,}
    \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.99}\NormalTok{, }\AttributeTok{max\_treedepth =} \DecValTok{15}\NormalTok{),}
    \AttributeTok{backend =} \StringTok{"cmdstanr"}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gp.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

And we plot the posterior predictions\ldots{}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# plotting the posterior predictions}
\FunctionTok{plot\_post\_preds}\NormalTok{(}\AttributeTok{model =}\NormalTok{ gam) }\SpecialCharTok{+} \FunctionTok{plot\_post\_preds}\NormalTok{(}\AttributeTok{model =}\NormalTok{ gp\_model)}
\end{Highlighting}
\end{Shaded}

\caption{Posterior predictions of the GAM (left) and GP (right) models.}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/gam-gp-preds-1.pdf}

@@

We can retrieve the posterior probability of the slope for
\texttt{condition} at each timestep (Figure~\ref{fig-plot-post-slope}).

@@

\caption{\label{fig-plot-post-slope}Slope for the difference between
conditions according to the GAM (left) or the GP (right).}

@@

We can also compute the posterior probability of the slope for
\texttt{condition} being above \(0+\epsilon\)
(Figure~\ref{fig-post-prob-test}), with \(\epsilon\) defined as 10\% of
the standard deviation of the raw EEG signal.

@@

\caption{\label{fig-post-prob-test}Posterior probability of the ERP
difference (slope) being above 0 according to the GAM (left) and the GP
(right).}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-post-prob-test-1.pdf}

@@

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# fitting the GP}
\NormalTok{meta\_gp }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{gp}\NormalTok{(time, }\AttributeTok{k =} \DecValTok{20}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{control =} \FunctionTok{list}\NormalTok{(}\AttributeTok{adapt\_delta =} \FloatTok{0.99}\NormalTok{),}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{4}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gp.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# plotting the posterior predictions}
\NormalTok{meta\_gam\_preds }\OtherTok{\textless{}{-}} \FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gam, }\AttributeTok{effect =} \StringTok{"time:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{scale\_colour\_met\_d}\NormalTok{(}\AttributeTok{name =} \StringTok{"Johnson"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{scale\_fill\_met\_d}\NormalTok{(}\AttributeTok{name =} \StringTok{"Johnson"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}
\NormalTok{meta\_gp\_preds }\OtherTok{\textless{}{-}} \FunctionTok{plot}\NormalTok{(}
    \FunctionTok{conditional\_effects}\NormalTok{(}\AttributeTok{x =}\NormalTok{ meta\_gp, }\AttributeTok{effect =} \StringTok{"time:condition"}\NormalTok{),}
    \AttributeTok{points =} \ConstantTok{FALSE}\NormalTok{, }\AttributeTok{theme =} \FunctionTok{theme\_light}\NormalTok{(), }\AttributeTok{plot =} \ConstantTok{FALSE}
\NormalTok{    )[[}\DecValTok{1}\NormalTok{]] }\SpecialCharTok{+}
    \FunctionTok{scale\_colour\_met\_d}\NormalTok{(}\AttributeTok{name =} \StringTok{"Johnson"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{scale\_fill\_met\_d}\NormalTok{(}\AttributeTok{name =} \StringTok{"Johnson"}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}\AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{, }\AttributeTok{y =} \StringTok{"EEG signal (a.u.)"}\NormalTok{)}

\CommentTok{\# combining the two plots}
\NormalTok{meta\_gam\_preds }\SpecialCharTok{+}\NormalTok{ meta\_gp\_preds}
\end{Highlighting}
\end{Shaded}

\caption{Posterior predictions for the hierarchical GAM and GP.}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/meta-gp-preds-1.pdf}

@@

\subsection{Application to actual M/EEG
data}\label{application-to-actual-meeg-data}

Assessing the reliability of the proposed approach (in comparison to
other methods) using some sort of split-half reliability
(\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018})?

@@

We considered five methods for multiplicity corrections: two FDR
methods; two cluster-based methods; and the maximum statistics. The two
FDR methods were BH95 (Benjamini \& Hochberg, 1995;
10.1111/j.2517-6161.1995.tb02031.x) and BY01 (Benjamini \& Yekutieli,
2001; 10.1214/aos/1013699998), which were applied to the permutation
p-values using the \texttt{p.adjust} function in \texttt{R}. The first
cluster-based inference was implemented using a cluster-sum statistic of
squared t-values\ldots{}

We also compared the performance of the GAMM (in correctly
identifying/estimating the onset and offset of ERP difference across
conditions) to the performance of the binary segmentation method as
implemented in the \texttt{R} package \texttt{changepoint} v 2.2.4
(\citeproc{ref-changepoint}{Killick et al., 2022a}), threshold-free
cluster enhancement (TFCE, \citeproc{ref-smith2009}{Smith \& Nichols,
2009}), as implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), and using the raw t-, p-,
or s-values timecourse. S-values
(\citeproc{ref-greenland2019}{Greenland, 2019})\ldots{}

\caption{Timecourse of squared t-values and s-values (continuous measure
of evidence given by -log2(p-value)) with true onset and onset
identified using the raw (uncorrected) p-values or the corrected
p-values (BH, BY, Holm).}

{\centering \includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/test-1d-1.pdf}

@@

\caption{T-values timecourse with true onset/offset and onset/offset
identified using the changepoint package (binary segmentation method, in
green).}

{\centering \includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/changepoint-1.pdf}

@@

Now cluster-based permutations using \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}) via the \texttt{R} package
\texttt{reticulate} v 1.35.0 (\citeproc{ref-reticulate}{Ushey et al.,
2024})\ldots{}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# importing the decoding scores}
\NormalTok{p\_values\_df }\OtherTok{\textless{}{-}} \FunctionTok{readRDS}\NormalTok{(}\AttributeTok{file =} \StringTok{"results/mne\_permutation\_decoding\_scores.rds"}\NormalTok{)}

\CommentTok{\# plotting the s{-}values ({-}log2(p{-}values))}
\NormalTok{p\_values\_df }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{sval =} \SpecialCharTok{{-}}\FunctionTok{log2}\NormalTok{(pval) ) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ggplot}\NormalTok{(}\FunctionTok{aes}\NormalTok{(}\AttributeTok{x =}\NormalTok{ time, }\AttributeTok{y =}\NormalTok{ sval) ) }\SpecialCharTok{+}
    \FunctionTok{geom\_line}\NormalTok{(}\AttributeTok{linewidth =} \FloatTok{0.5}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{geom\_hline}\NormalTok{(}\AttributeTok{yintercept =} \SpecialCharTok{{-}}\FunctionTok{log2}\NormalTok{(}\FloatTok{0.05}\NormalTok{), }\AttributeTok{linetype =} \DecValTok{2}\NormalTok{) }\SpecialCharTok{+}
    \FunctionTok{labs}\NormalTok{(}
        \AttributeTok{x =} \StringTok{"Time (s)"}\NormalTok{,}
        \AttributeTok{y =} \StringTok{"{-}log2(p{-}value)"}
\NormalTok{        )}
\end{Highlighting}
\end{Shaded}

@@

\caption{\label{fig-mne-cluster}Cluster-based permutation tests via
MNE-Python.}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-mne-cluster-1.pdf}

@@

>>>>

\title{Modelling M/EEG data with Bayesian generalised additive
multilevel models}

@@

\abstract{Time-resolved electrophysiological measurements such as those
offered by magneto- or electro-encephalography (M/EEG) provide a unique
window onto neural activity underlying cognitive process and how they
unfold over time. Typically, we are interested in testing whether such
measures differ across conditions and/or groups. The conventional
approach consists in conducting mass-univariate statistics through
followed by some form of multiplicity correction (e.g., FDR, FWER) or
cluster-based inference. However, these cluster-based methods have an
important downside: they shift the focus of inference from the timepoint
to the cluster level, thus preventing any conclusion to be made about
the onset and offset of effects (e.g., differences across conditions).
Here, we introduce a novel \emph{model-based approch} for analysing
one-dimensional M/EEG timeseries such as ERPs or decoding timecourses
and their differences across conditions or group. This approach relies
on Bayesian generalised additive multilevel models, which output the
posterior probabilility of the effect being above 0 (or above chance) at
every timestep, while naturally taking into account the temporal
dependencies and between-subject variability present in such data.}

\keywords{EEG, MEG, generalised additive models, mixed-effects
models, multilevel models, Bayesian statistics, brms}

@@

\hypertarget{toc}{}
\tableofcontents
\newpage
\section[Introduction]{Modelling M/EEG data with Bayesian generalised
additive multilevel models}

@@

\section{Introduction}\label{introduction}

Here are some useful references to be discussed
(\citeproc{ref-combrisson_exceeding_2015}{Combrisson \& Jerbi, 2015};
\citeproc{ref-ehinger_unfold_2019}{Ehinger \& Dimigen, 2019};
\citeproc{ref-frossard2021}{Frossard \& Renaud, 2021},
\citeproc{ref-frossard2022}{2022}; \citeproc{ref-gramfort2013}{Gramfort,
2013}; \citeproc{ref-hayasaka_validating_2003}{Hayasaka, 2003};
\citeproc{ref-luck_how_2017}{Luck \& Gaspelin, 2017};
\citeproc{ref-maris2007}{Maris \& Oostenveld, 2007};
\citeproc{ref-pedersen_hierarchical_2019}{E. J. Pedersen et al., 2019};
\citeproc{ref-pernet2015}{C. R. Pernet et al., 2015})\ldots{} See also
(\citeproc{ref-maris2011}{Maris, 2011})\ldots{} and
(\citeproc{ref-rosenblatt2018}{Rosenblatt et al., 2018}) (history of
cluster-based approaches and using a data split?)\ldots{} Cluster
failure (\citeproc{ref-eklund2016}{Eklund et al., 2016})\ldots{}

@@

@@

From Dimigen \& Ehinger (\citeproc{ref-dimigen2021}{2021}): ``Recently,
spline regression has been applied to ERPs (Hendrix, Baayen, \& Bolger,
2017; Kryuchkova,Tucker, Wurm, \& Baayen,2012; Tremblay \& Baayen, 2010;
Tremblay \& Newman, 2015)''\ldots{} GAMMs for EEG data
(\citeproc{ref-abugaber2023}{Abugaber et al., 2023};
\citeproc{ref-meulman2015}{Meulman et al., 2015})\ldots{}

Disentangling overlapping processes
(\citeproc{ref-skukies_brain_2024}{Skukies et al., 2024};
\citeproc{ref-skukies_modelling_2021}{Skukies \& Ehinger, 2021})\ldots{}
Weighting single trials (\citeproc{ref-pernet2022}{C. Pernet,
2022})\ldots{} The LIMO toolbox
(https://onlinelibrary.wiley.com/doi/10.1155/2011/831409)\ldots{} Using
Bayes factors (\citeproc{ref-teichmann2022}{Teichmann, 2022})\ldots{}

@@

See for instance these tutorials
(\citeproc{ref-suxf3skuthy2017}{Sóskuthy, 2017};
\citeproc{ref-winter2016}{Winter \& Wieling, 2016}) or application to
phonetic data (\citeproc{ref-wieling2018}{Wieling, 2018}) or this
introduction (\citeproc{ref-baayen2020}{Baayen \& Linke, 2020}) or these
reference books (\citeproc{ref-hastie2017}{Hastie \& Tibshirani, 2017};
\citeproc{ref-wood2017}{Wood, 2017})\ldots{}

@@

Focusing on identifying onset and offset of effects (as assessed by ERP
differences or decoding performance)\ldots{} Assessing the performance
of a model-based approach (i.e., Bayesian GAMMs) to conventional methods
(multiplicity corrections or cluster-based permutation)\ldots{}

@@

We then fitted a Bayesian GAM using the \texttt{brms} package
(\citeproc{ref-brms2017}{Bürkner, 2017}, \citeproc{ref-brms2018}{2018};
\citeproc{ref-nalborczyk2019}{Nalborczyk et al., 2019}). We used the
default priors in \texttt{brms}, that is, weakly informative priors. We
ran eight Markov Chain Monte-Carlo (MCMC) to approximate the posterior
distribution, including each 5000 iterations and a warmup of 2000
iterations, yielding a total of \(8 \times (5000-2000) = 24000\)
posterior samples to use for inference. Posterior convergence was
assessed examining trace plots as well as the Gelman--Rubin statistic
\(\hat{R}\).

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# averaging across participants}
\NormalTok{ppt\_df }\OtherTok{\textless{}{-}}\NormalTok{ raw\_df }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{group\_by}\NormalTok{(participant, condition, time) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{summarise}\NormalTok{(}\AttributeTok{eeg =} \FunctionTok{mean}\NormalTok{(eeg) ) }\SpecialCharTok{\%\textgreater{}\%}
    \FunctionTok{ungroup}\NormalTok{()}

@@

\CommentTok{\# fitting the GAM}
\NormalTok{gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# cubic regression splines with k{-}1 basis functions}
\NormalTok{    eeg }\SpecialCharTok{\textasciitilde{}}\NormalTok{ condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition),}
    \AttributeTok{data =}\NormalTok{ ppt\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

We then plot the posterior predictions together with the posterior
estimate of the slope for \texttt{condition} at each timestep
(Figure~\ref{fig-plot-post-slope}).

@@

\caption{\label{fig-plot-post-slope}Posterior estimate of the ERP in
each condition (left) or directly for the difference of ERPs (right)
according to the GAM.}

@@

We then compute the posterior probability of the slope for
\texttt{condition} being above \(0+\epsilon\)
(Figure~\ref{fig-post-prob-test}), with \(\epsilon\) defined as 10\% of
the standard deviation of the raw EEG signal.

@@

\caption{\label{fig-post-prob-test}Posterior probability of the ERP
difference (slope) being above 0 according to the GAM.}

@@

\includegraphics[width=0.75\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-post-prob-test-1.pdf}

@@

\CommentTok{\# fitting the GAM}
\NormalTok{meta\_gam }\OtherTok{\textless{}{-}} \FunctionTok{brm}\NormalTok{(}
    \CommentTok{\# using by{-}participant SD of ERPs across trials}
\NormalTok{    eeg\_mean }\SpecialCharTok{|} \FunctionTok{se}\NormalTok{(eeg\_sd) }\SpecialCharTok{\textasciitilde{}}
\NormalTok{        condition }\SpecialCharTok{+} \FunctionTok{s}\NormalTok{(time, }\AttributeTok{bs =} \StringTok{"cr"}\NormalTok{, }\AttributeTok{k =} \DecValTok{10}\NormalTok{, }\AttributeTok{by =}\NormalTok{ condition) }\SpecialCharTok{+}
\NormalTok{        (}\DecValTok{1} \SpecialCharTok{|}\NormalTok{ participant),}
    \AttributeTok{data =}\NormalTok{ summary\_df,}
    \AttributeTok{family =} \FunctionTok{gaussian}\NormalTok{(),}
    \AttributeTok{warmup =} \DecValTok{2000}\NormalTok{,}
    \AttributeTok{iter =} \DecValTok{5000}\NormalTok{,}
    \AttributeTok{chains =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{cores =} \DecValTok{8}\NormalTok{,}
    \AttributeTok{file =} \StringTok{"models/meta\_gam.rds"}
\NormalTok{    )}
\end{Highlighting}
\end{Shaded}

@@

@@

We compared the ability of the GAMM to correctly estimate the onset and
offset of the ERP difference to widely used methods. First, we conducted
mass-univariate t-tests (thus treating each timestep independently) and
identified the onset and offset of the ERP difference as the first and
last values crossing an arbitrary significance threshold
(\(\alpha = 0.01\)). We then followed the same approach but after
applying different forms of multiplicity correction the the p-values. We
compared two methods that control the false discovery rate (FDR) (i.e.,
BH95, \citeproc{ref-benjamini1995}{Benjamini \& Hochberg, 1995}; and
BY01, \citeproc{ref-benjamini2001}{Benjamini \& Yekutieli, 2001}), one
method that controls the familywise error rate (FWER) (i.e.,
Holm--Bonferroni method, https://www.jstor.org/stable/4615733), and two
cluster-based methods (permutation with a single threshold and TFCE,
\citeproc{ref-smith2009}{Smith \& Nichols, 2009}). The BH95, BY01, and
Holm corrections were applied to the p-values using the
\texttt{p.adjust()} function in \texttt{R}. The cluster-based inference
was implemented using a cluster-sum statistic of squared t-values, as
implemented in \texttt{MNE-Python}
(\citeproc{ref-gramfort2013}{Gramfort, 2013}), called via the \texttt{R}
package \texttt{reticulate} v 1.35.0 (\citeproc{ref-reticulate}{Ushey et
al., 2024}). We also compared these estimates to the onset and offset as
estimated using the binary segmentation algorithm, as implemented in the
\texttt{R} package \texttt{changepoint} v 2.2.4
(\citeproc{ref-changepoint}{Killick et al., 2022a}), and applied
directly to the squared t-values (as in
\citeproc{ref-rousselet_using_2025}{Rousselet, 2025}). For visualisation
and interpretability purposes, we converted p-values to s-values, which
can be interpreted as bits of surprising information, assuming a null
effect (\citeproc{ref-greenland2019}{Greenland, 2019})
(Figure~\ref{fig-corrections}).

\caption{\label{fig-corrections}Timecourse of squared t-values and
s-values, with true onset (black dashed line) and onsets identified
using the raw (uncorrected) p-values or the corrected p-values (BH, BY,
Holm).}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-corrections-1.pdf}

@@

Onset/offset estimation methods were assessed using these summary
statistics of the Monte Carlo sampling distributions of onsets: median
absolute error (MAE) and variance, with 10.000 iterations (i.e., 10.000
simulated datasets)\ldots{}

\subsection{Application to actual MEG
data}\label{application-to-actual-meg-data}

Assessing the reliability of the proposed approach using some sort of
split-half reliability (e.g., \citeproc{ref-rosenblatt2018}{Rosenblatt
et al., 2018})? Using the MEG decoding results of Nalborczyk et al.~(in
preparation) in 33 participants:

\begin{itemize}
\tightlist
\item
  Create many (e.g., 10 or 20) half train/test splits of the data
\item
  For each fold, estimate the onset/offset on both splits using all
  methods
\item
  Then summarise the distribution of between-split difference across
  folds
\end{itemize}

This will allow checking that the proposed approach produces more
reliable onset/offset estimates.

\subsection{Simulation results (bias and
variance)}\label{simulation-results-bias-and-variance}

Figure~\ref{fig-simulation-mae-variance} shows a summary of the
simulation results, revealing that the proposed approach (\texttt{brms})
has the lowest MAE and variance for both the onset and offset
estimates\ldots{}

\caption{\label{fig-simulation-mae-variance}Median absolute error and
variance of onset and offset estimates for each method.}

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-simulation-mae-variance-1.pdf}

@@

@@

\caption{\label{fig-simulation-distribution}Distributions of onset and
offset estimates for each method.}

@@

\includegraphics[width=1\textwidth,height=\textheight]{brms_meeg_files/figure-pdf/fig-simulation-distribution-1.pdf}

@@

\section{Data and code availability}\label{data-and-code-availability}
